---
title: "Lottery Analysis"
author: sebastian deri
date: jan-feb 2021
output: html_notebook
---

# --- Settings ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
plot_save <- FALSE
```


# --- Packages ---

```{r}
library(tidyverse)
library(patchwork)

```


# --- Data ---

# 1. create data

```{r}
# make vector of possible incomes
money <- seq(0, 100, 1)

# turn into data frame
df_util <- data.frame(money)

# show df
df_util

```

# --- Analysis ---

# 2. linear utility

## 2.1. make data

```{r}
max_income <- 200e3
income <- seq(0, 1, 0.01) * max_income
happiness <- income/max(income) * 1000

df_linear <- data.frame(income = income,
                        happiness = happiness)

df_linear
```

## 2.2. actually graph

```{r, fig.width=10, fig.height=6}
plot_linear <- 
df_linear %>% 
  ggplot(aes(x = income,
             y = happiness)) +
  geom_line() +
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness") +
  scale_y_continuous(breaks = seq(0, max(df_linear$happiness), 100),
                     labels = scales::comma) +
  scale_x_continuous(breaks = seq(0, max_income, 25e3),
                     labels = scales::dollar_format()) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))
plot_linear

```

# 3. concave utility

## 3.1. make data

```{r}
max_income <- 200e3
income <- seq(0, 1, 0.01) * max_income
happiness <- (1-(abs(income-max_income)^2.5)/(max_income^2.5)) * 1000

df_concave <- data.frame(income = income,
                        happiness = happiness)

df_concave
```

## 3.2. actually graph

```{r, fig.width=10, fig.height=6}
plot_concave <-
df_concave %>% 
  ggplot(aes(x = income,
             y = happiness)) +
  geom_line() +
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness \n (Measured in Personal Happiness Points)") +
  scale_y_continuous(breaks = seq(0, max(df_concave$happiness), 100),
                     labels = scales::comma) +
  #scale_y_continuous(limits = c(0,  1))
  scale_x_continuous(breaks = seq(0, max_income, 25e3),
                     labels = scales::dollar_format()) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))
plot_concave

```

# 4. convex utility

## 4.1. make data

going to cheat here and kind of (try to) flip the concave utility function over a 45 degree diagonal line

```{r}
# linear utility
max_income <- 200e3
income <- seq(0, 1, 0.01) * max_income
happiness <- income/max(income) * 1000

df_linear <- data.frame(income = income,
                        happiness = happiness)

df_linear

# concave utility
max_income <- 200e3
happiness <- (1-(abs(income-max_income)^2.5)/(max_income^2.5)) * 1000

df_concave <- data.frame(income = income,
                        happiness = happiness)

df_concave

# convex utility
df_convex <- data.frame(income = income,
                        neg_income = -income+max_income,
                        happiness = df_concave$happiness,
                        happiness_trans1 = -df_concave$happiness + 1000)

df_convex

```
## 4.2. test graph (all 4 quadrants)

```{r, fig.width=10, fig.height=6}
df_convex %>% 
  ggplot() +
  geom_line(aes(x = income,
                y = happiness),
            color = "black") +
  geom_line(aes(x = neg_income,
                y = happiness_trans1),
            color = "red") +
  #geom_line(aes(y = happiness_trans2),
  #          color = "blue") +
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness \n (Measured in Personal Happiness Points)") +
  scale_y_continuous(breaks = seq(-max(abs(df_convex$happiness)), max(abs(df_convex$happiness)), 100),
                     limits = c(-max(abs(df_convex$happiness)), max(abs(df_convex$happiness))),
                     labels = scales::comma) +
  #scale_y_continuous(limits = c(0,  1))
  scale_x_continuous(breaks = seq(-max_income, max_income, 25e3),
                     limits = c(-max_income, max_income),
                     labels = scales::dollar_format()) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```

## 4.3. actually graph

```{r, fig.width=10, fig.height=6}
plot_convex <- 
df_convex %>% 
  ggplot(aes(x = neg_income,
             y = happiness_trans1)) +
  geom_line() +
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness \n (Measured in Personal Happiness Points)") +
  scale_y_continuous(breaks = seq(0, max(df_convex$happiness), 100),
                     labels = scales::comma) +
  #scale_y_continuous(limits = c(0,  1))
  scale_x_continuous(breaks = seq(0, max_income, 25e3),
                     labels = scales::dollar_format()) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))
plot_convex

```

# 5. all plots together

## 5.1. old way

```{r, fig.width=10, fig.height=6}
plot_linear + plot_concave + plot_convex
```

## 5.1. new way

```{r, fig.width=15, fig.height=6}
(plot_linear | plot_concave | plot_convex)
```


# 6. Plots w/ Shifts

## 6.1. linear

```{r, fig.width=10, fig.height=6}
# presets
mid_x <- 100e3
mid_y <- df_linear[df_linear$income == mid_x,]$happiness
mid_color <- "gray26"

low_x <- 50e3
low_y <- df_linear[df_linear$income == low_x,]$happiness
low_color <- "red3"
low_shift <- 5e3

high_x <- 150e3
high_y <- df_linear[df_linear$income == high_x,]$happiness
high_color <- "green4"
high_shift <- 5e3


df_linear %>% 
  ggplot(aes(x = income,
             y = happiness)) +
  geom_line() +
  # middle
  geom_segment(aes(x = mid_x, xend = mid_x, y = 0, yend = mid_y),
               color = mid_color, linetype = "dotted", size = 1) +
  geom_segment(aes(x = 0, xend = mid_x, y = mid_y, yend = mid_y),
               color = mid_color, linetype = "dotted", size = 1) +
  # low
  geom_segment(aes(x = low_x, xend = low_x, y = 0, yend = low_y),
               color = low_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = 0, xend = low_x, y = low_y, yend = low_y),
               color = low_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = mid_x - low_shift, xend = low_x + low_shift , y = low_y/2, yend = low_y/2),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = low_color) +
  annotate(geom = "text", x = low_x + (mid_x-low_x)/2, y = low_y - 10, label = "Lose $50K",
           angle = 0, size = 3.5, color = low_color) +
  geom_segment(aes(x = -10e3, xend = -10e3 , y = mid_y - 15, yend = low_y + 15),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = low_color) +
  annotate(geom = "text", x = 25e3, y = low_y + (mid_y-low_y)/2, label = "-100 Happiness",
           angle = 0, size = 3.5, color = low_color) +
  # high
  geom_segment(aes(x = high_x, xend = high_x, y = 0, yend = high_y),
               color = high_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = 0, xend = high_x, y = high_y, yend = high_y),
               color = high_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = mid_x + high_shift, xend = high_x - low_shift , y = low_y/2, yend = low_y/2),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = high_color) +
  annotate(geom = "text", x = high_x - (high_x-mid_x)/2, y = low_y - 10, label = "Gain $50K",
           angle = 0, size = 3.5, color = high_color) +
  geom_segment(aes(x = -10e3, xend = -10e3 , y = mid_y + 15, yend = high_y - 15),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = high_color) +
  annotate(geom = "text", x = 25e3, y = high_y - (high_y - mid_y)/2, label = "+100 Happiness",
           angle = 0, size = 3.5, color = high_color) +
  # naming & formatting
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness") +
  scale_y_continuous(breaks = seq(0, max(df_linear$happiness), 100),
                     labels = scales::comma) +
  scale_x_continuous(breaks = seq(0, max_income, 25e3),
                     labels = scales::dollar_format()) +
  coord_cartesian(xlim = c(0, max(df_linear$income))) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```

## 6.2. concave

```{r, fig.width=10, fig.height=6}
# presets
mid_x <- 100e3
mid_y <- (1-(abs(mid_x-max_income)^2.5)/(max_income^2.5)) * 1000  # NOTE: max_income set elsewhere
mid_color <- "gray26"

low_x <- 50e3
low_y <- (1-(abs(low_x-max_income)^2.5)/(max_income^2.5)) * 1000  # NOTE: max_income set elsewhere
low_color <- "red3"
low_shift <- 5e3

high_x <- 150e3
high_y <- (1-(abs(high_x-max_income)^2.5)/(max_income^2.5)) * 1000  # NOTE: max_income set elsewhere
high_color <- "green4"
high_shift <- 5e3


df_concave %>% 
  ggplot(aes(x = income,
             y = happiness)) +
  geom_line() +
  # middle
  geom_segment(aes(x = mid_x, xend = mid_x, y = 0, yend = mid_y),
               color = mid_color, linetype = "dotted", size = 1) +
  geom_segment(aes(x = 0, xend = mid_x, y = mid_y, yend = mid_y),
               color = mid_color, linetype = "dotted", size = 1) +
  # low
  geom_segment(aes(x = low_x, xend = low_x, y = 0, yend = low_y),
               color = low_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = 0, xend = low_x, y = low_y, yend = low_y),
               color = low_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = mid_x - low_shift, xend = low_x + low_shift , y = low_y/2, yend = low_y/2),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = low_color) +
  annotate(geom = "text", x = low_x + (mid_x-low_x)/2, y = low_y - 70, label = "Lose $50K",
           angle = 0, size = 3.5, color = low_color) +
  geom_segment(aes(x = -10e3, xend = -10e3 , y = mid_y - 15, yend = low_y + 15),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = low_color) +
  annotate(geom = "text", x = 25e3, y = low_y + (mid_y-low_y)/2, label = paste0("-", round(mid_y-low_y, 0)," Happiness"),
           angle = 0, size = 3.5, color = low_color) +
  # high
  geom_segment(aes(x = high_x, xend = high_x, y = 0, yend = high_y),
               color = high_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = 0, xend = high_x, y = high_y, yend = high_y),
               color = high_color, linetype = "solid", size = 1) +
  geom_segment(aes(x = mid_x + high_shift, xend = high_x - low_shift , y = low_y/2, yend = low_y/2),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = high_color) +
  annotate(geom = "text", x = high_x - (high_x-mid_x)/2, y = low_y - 70, label = "Gain $50K",
           angle = 0, size = 3.5, color = high_color) +
  geom_segment(aes(x = -10e3, xend = -10e3 , y = mid_y + 15, yend = high_y - 15),
                  arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = high_color) +
  annotate(geom = "text", x = 25e3, y = high_y - (high_y - mid_y)/2, label = paste0("+", round(high_y-mid_y, 0)," Happiness"),
           angle = 0, size = 3.5, color = high_color) +
  # naming & formatting
  labs(title = "How Money Translates to Happiness",
       x = "Total Wealth ($)",
       y = "Total Happiness") +
  scale_y_continuous(breaks = seq(0, max(df_concave$happiness), 100),
                     labels = scales::comma) +
  #scale_y_continuous(limits = c(0,  1))
  scale_x_continuous(breaks = seq(0, max_income, 25e3),
                     labels = scales::dollar_format()) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10, b = 0, r = 0, l = 0)),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))
```


# END/MISC


# --- END/MISC ---

